Re-examining cross-cultural similarity judgments using lexical co-occurrence

### 2023-02-01 code/4_paper/manuscript/manuscript.Rmd
manuscript edits

Many cognitive processes rely on similarity, from inference and generalization to analogy, mathematics, and science. There is substantial consistency in human similarity reasoning, but also systematic variation across cultural and linguistic contexts. In particular, there is considerable evidence showing that similarity reasoning in East Asian cultural contexts differs from that in Western cultures (e.g. @Nisbett2003b).
<!-- Khuyen could you add this ref as Nisbett2003b?  https://www.pnas.org/doi/10.1073/pnas.1934527100-->
For example, in a triad task comparing preferences for taxonomic and thematic similarity (choose two out of three words that are most related to one another), @Ji2004 found that Chinese participants preferred thematic matching to a greater extent than European Americans.
In an image version of this task, Chinese children (9-10 years old) are also more likely to choose thematic matches compared to their American counterparts [@Chiu1972].
This cross-cultural difference is also observed in novel object categorization, with Chinese participants preferring to group by family resemblance across multiple features and Americans preferring a single-feature rule [@Norenzayan2002].
Across tasks, East Asian participants show a preference for thematic similarity based on causal, spatial, and temporal relationships while Western participants are more likely to make taxonomic matches based on the similarity of attributes, like shared color or shape, among objects [see @Markman1984 for a discussion of these types of similarity].

@@

To determine whether varied input to similarity judgments can in part explain cross-cultural differences, we need a way to operationalize variation in exposure to thematic and taxonomic relations.
While exposure to types of similarity is difficult to measure, the statistics of language can provide a rough proxy.
Language statistics are useful in that they are part of the input to everyday experience -- and indeed, may afford many of the 'experiences' that people have with infrequently encountered items, like cows or helicopters -- and they provide an accessible measure.
Previous work also suggests that language statistics, such as lexical co-occurrence or cosine distance of word embeddings, can be good predictors of similarity reasoning.
Semantic models that are constructed using lexical co-occurrence (in comparison to annotated relations) have been shown to perform well on predicting human judgments about similarity between word pairs that are thematically or taxonomically related [@Rohde2006]. 
Relatedly, a model trained on word-document co-occurrence can predict word association and the effects of semantic association on a variety of linguistic tasks [@Griffiths2007].
Word embeddings like word2vec, gloVe, and fastText have also been shown to be good predictors for similarity judgments (@Liu2019, @Jatnika2019).

>>>>

Many cognitive processes rely on similarity, from inference and generalization to analogy, mathematics, and science. There is substantial consistency in human similarity reasoning, but also systematic variation across cultural and linguistic contexts. In particular, there is considerable evidence showing that similarity reasoning in East Asian cultural contexts differs from that in Western cultures [e.g. @Nisbett2003b].
For example, in a triad task comparing preferences for taxonomic and thematic similarity (choose two out of three words that are most related to one another), @Ji2004 found that Chinese participants preferred thematic matching to a greater extent than European Americans.
In an image version of this task, Chinese children (9-10 years old) are also more likely to choose thematic matches compared to their American counterparts [@Chiu1972].
This cross-cultural difference is also observed in novel object categorization, with Chinese participants preferring to group by family resemblance across multiple features and Americans preferring a single-feature rule [@Norenzayan2002].
Across tasks, East Asian participants show a preference for thematic similarity based on causal, spatial, and temporal relationships while Western participants are more likely to make taxonomic matches based on the similarity of attributes, like shared color or shape, among objects [see @Markman1984 for a discussion of these types of similarity].

@@

To determine whether varied input to similarity judgments can in part explain cross-cultural differences, we need a way to operationalize variation in exposure to thematic and taxonomic relations.
While exposure to types of similarity is difficult to measure, the statistics of language can provide a rough proxy.
Language statistics are useful in that they are part of the input to everyday experience -- and indeed, may afford many of the 'experiences' that people have with infrequently encountered items, like cows or helicopters -- and they provide an accessible measure.
Previous work also suggests that language statistics, such as lexical co-occurrence or cosine distance of word embeddings, can be good predictors of similarity reasoning.
Semantic models that are constructed using lexical co-occurrence (in comparison to annotated relations) have been shown to perform well on predicting human judgments about similarity between word pairs that are thematically or taxonomically related [@Rohde2006]. 
Relatedly, a model trained on word-document co-occurrence can predict word association and the effects of semantic association on a variety of linguistic tasks [@Griffiths2007].
Word embeddings like word2vec, gloVe, and fastText have also been shown to be good predictors for similarity judgments [@Liu2019, @Jatnika2019].

### 2023-02-01 code/4_paper/manuscript/manuscript.tex
manuscript edits

\textbf{Keywords:}
similarity; culture; language; semantics; lexical co-occurrence;
variation
\end{abstract}

@@

We constantly partition our continuous experience of the world into
categories, but the way we organize these partitions can vary across
cultures. One way this cross-cultural difference has been observed is
with a task comparing preference for taxonomic and thematic similarity
between East Asian and Western participants.\footnote{Taxonomic
  categorization is based on the similarity of attributes, for example,
  similar perceptual properties, like shared color or shape, among
  objects.In contrast, thematic categorization is based on causal,
  spatial, and temporal relationships among objects (Markman \&
  Hutchinson, 1984).} Ji, Zhang, \& Nisbett (2004) found that in a word
triad task (choose two out of three words that are more related to one
another), Chinese participants preferred thematic matching more compared
to European Americans. In a pictorial version of this task, Chiu (1972)
found that Chinese children (9-10 years old) are also more likely to
choose thematic matches compared to their American counterparts. This
cross-cultural difference is also observed in novel object
categorization, with Chinese participants preferring to group by family
resemblance across multiple features and Americans preferring a
single-feature rule (Norenzayan, Smith, Kim, \& Nisbett, 2002).

One hypothesis is that differences in similarity judgment is linked to
the tendency to see more analytic processing in Western culture and more
holistic processing in East Asian cultures (Nisbett, 2003). On this
hypothesis, since analytic processing emphasizes relationships between
objects and their properties, it leads to taxonomic similarity
judgments; in contrast, holistic processing emphasizes relations between
objects and leads to thematic similarity judgments. Cross-cultural
difference in similarity judgment is thus connected to work showing East
Asian participants exhibiting a higher level of sensitivity to context
than their Western counterparts when reproducing drawings from memory
(Ji, Peng, \& Nisbett, 2000); visually exploring naturalistic scenes
(Chua, Boland, \& Nisbett, 2005); describing scenes (Masuda \& Nisbett,
2001); and in explaining the causes of ambiguous behaviors (Choi,
Nisbett, \& Norenzayan, 1999). This view ascribes cross-cultural
differences in similarity judgment to variation in how each culture (or
group of culture, i.e.~East Asian vs Western) \emph{conceptualize}
similarity -- with East Asians preferring thematic relations more than
Westerners.

Alternatively, these judgments could be shaped by cross-cultural
differences in the \emph{inputs} to similarity judgment, namely, the
statistics of the environment, and accordingly the content of everyday
experiences. Perhaps when faced with the triad task, participants from
all cultures follow the same process for conceptualizing similarity, but
rely on language or culture-specific inputs to this process. If we
observe a difference in categorization between East Asian and Western
participants, it could be that members of both groups use the same
procedure (considering similarity that is influenced by both taxonomic
and thematic relations), but the inputs to this procedure differ between
cultures, with East Asian participants exposed to more support for
thematic similarity in the language they hear in comparison to their
Western counterparts. We might also expect that both conceptualization
and inputs of similarity play a role in driving cross-cultural
differences -- East Asian participants may both be exposed to more
instances of thematic similarity, and prefer to conceptualize similarity
as thematic relations more so than Westerners.

@@

To test alternative hypotheses involving variation of inputs to
similarity judgment, we need a way to operationalize variation in
exposure to thematic vs taxonomic relations. While this metric is
difficult to directly measure, language statistics can provide a rough
proxy. Language statistics is useful in that it is part of the inputs to
everyday experiences (through spoken and written language), may afford
many of the `experiences' that people have with infrequently encountered
items, like cows or helicopters, and provides accessible measures.
Previous work also suggests that using language statistics (such as
lexical co-occurrence or cosine distance of word embeddings) as a proxy
can be a good predictor for similarity reasoning. A model trained on
word-document co-occurrence can predict word association and the effects
of semantic association on a variety of linguistic tasks (Griffiths,
Steyvers, \& Tenenbaum, 2007). Lexical semantic models that are
constructed using lexical co-occurrence (in comparison to annotated
relations) have been shown to perform well on predicting human judgments
about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006). Word
embeddings like word2vec, gloVe, and fastText have been demonstrated to
be good predictors for similarity judgments (Jatnika, Bijaksana, \&
Suryani, 2019; Liu, Feng, Wu, Chan, \& Fulton, 2019).

Our study uses cosine distances of fastText word vectors as a measure of
lexical co-occurrence\footnote{We also carried out our analysis using
  raw lexical co-occurrences and obtained similar results.}. fastText is
a system that uses lexical co-occurrence information to generate a
vector representing each word in its lexicon (Mikolov, Grave,
Bojanowski, Puhrsch, \& Joulin, 2018). fastText has also been shown to
be sensitive towards cultural effects on word meanings: Thompson \&
Lupyan (2020) showed that the distribution of semantic meaning clusters
generated by fastText trained on language-specific corpora correlates to
the cultural, historical and geographical distances of such languages
languages. This is a proof of concept for our use of fastText in
different languages with varying levels of relatedness.

We note that language statistics can incorporate both cultural-specific
environmental statistics (the inputs of similarity judgment) and
thematic/taxonomic preferences (the conceptualization of similarity).
For example, cultural features (farming) can lead to differences in
environmental statistics (seeing cows and grass) and in turn these can
influence language (talking more about cows and grass). But cultural
features (preferring thematic relations) could also more directly cause
individuals to talk differently about the same experiences (mentioning
what cows eat rather than what other animals cows are like). Therefore,
our approach looks at the extent to which language statistics can
predict cross-cultural differences in similarity judgment with the
understanding that language statistics is a proxy for both inputs and
conceptualization of similarity judgment\footnote{However, in Q3, we
  investigate whether differing culture-based conceptualization of
  similarity has unique contribution to cross-cultural variation of
  similarity jdugment.}.

@@

Our study tests the hypothesis that, rather than (or in addition to) the
variation in conceptualization of similarity across cultural groups,
variation in the input to similarity judgment (namely, environmental
statistics as proxied by language statistics) can be used to predict
cross-cultural differences observed in how people evaluate similarity.

We measured taxonomic versus thematic categorization in a word triad
task in three populations. Following Ji et al. (2004), we conducted our
study in the US and China. In addition, we collected data in Vietnam,
Vietnam is a Southeast Asian country that borders China and is
historically greatly influenced by Chinese culture (Hui, 2002).
Therefore, it serves as a suitable cultural context to investigate
whether the claim made by Ji et al. (2004) and previous studies, that
Eastern and Western cultures have different notions of similarity,
extends beyond mainland China. We then tested if fastText vectors from
the corresponding corpus (English, Mandarin, and Vietnamese) are good
predictors for similarity judgment in each cultural context.

We note that our study is correlational and cannot make a causal claim
about the relationship of environmental statistics to similarity
judgment, but it may be possible to gain traction on potential
mechanisms by examining whether variation in similarity judgments covary
with environmental statistics that differ across cultural and linguistic
contexts. Our specific research questions are as follows:~

Q1: Do we replicate cross-cultural differences in thematic/taxonomic
categorization between East Asian and Western cultures?

Q2: Are these cross-cultural differences related to differences in
lexical co-occurrence?

Q3: Is there evidence for a unique contribution from conceptualization
of similarity?

Q4: Is the relationship with co-occurrence specific to
thematic/taxonomic items, or can co-occurrence structure predict other
judgments?~

To preview our results, we replicate US-China differences and find that
Vietnamese judgments are intermediate between these two. Further,
language-specific statistics are a good predictor for cross-cultural
differences in similarity judgments. Language statistics are also a good
predictor for filler items with no thematic/taxonomic structure. Our
findings are in contrast to previous accounts, by which culture induces
differing conceptions of similarity (varying between taxonomic and
thematic). these findings provide an alternative account by which
language may explain these cross-cultural differences without invoking
variation in notions of similarity.

@@

>>>>

\textbf{Keywords:}
similarity; culture; language; semantics; lexical co-occurrence;
variation; US; China; Vietnam
\end{abstract}

@@

Many cognitive processes rely on similarity, from inference and
generalization to analogy, mathematics, and science. There is
substantial consistency in human similarity reasoning, but also
systematic variation across cultural and linguistic contexts. In
particular, there is considerable evidence showing that similarity
reasoning in East Asian cultural contexts differs from that in Western
cultures (e.g. Nisbett \& Masada, 2003). For example, in a triad task
comparing preferences for taxonomic and thematic similarity (choose two
out of three words that are most related to one another), Ji, Zhang, \&
Nisbett (2004) found that Chinese participants preferred thematic
matching to a greater extent than European Americans. In an image
version of this task, Chinese children (9-10 years old) are also more
likely to choose thematic matches compared to their American
counterparts (Chiu, 1972). This cross-cultural difference is also
observed in novel object categorization, with Chinese participants
preferring to group by family resemblance across multiple features and
Americans preferring a single-feature rule (Norenzayan, Smith, Kim, \&
Nisbett, 2002). Across tasks, East Asian participants show a preference
for thematic similarity based on causal, spatial, and temporal
relationships while Western participants are more likely to make
taxonomic matches based on the similarity of attributes, like shared
color or shape, among objects (see Markman \& Hutchinson, 1984 for a
discussion of these types of similarity).

An influential perspective within cultural psychology links these
differences in similarity judgment to tendencies toward analytic
processing in Western cultures, and holistic processing in East Asian
cultures (Nisbett, 2003). Analytic processing emphasizes rule-like
relationships predicated on objects and their properties and,
correspondingly, taxonomic similarity, while holistic processing
emphasizes relations between objects and their context and therefor,
thematic similarity. In related work, East Asian participants show a
higher level of sensitivity to context than their Western counterparts
when reproducing drawings from memory (Ji, Peng, \& Nisbett, 2000);
visually exploring naturalistic scenes (Chua, Boland, \& Nisbett, 2005);
describing scenes (Masuda \& Nisbett, 2001); and in explaining the
causes of ambiguous behaviors (Choi, Nisbett, \& Norenzayan, 1999). The
consensus interpretation in this literature ascribes cross-cultural
differences in similarity judgment to variation in the
\emph{conceptualization} of similarity -- with people from East Asian
cultures relying on a more thematic notion of similarity than
Westerners.

Alternatively, these judgments could be shaped by cross-cultural
differences in the \emph{input} to similarity judgment, that is, the
statistics of the environment, and the content of everyday experiences.
Perhaps when faced with the triad task, participants from all cultures
follow the same process for conceptualizing similarity, but rely on
language or culture-specific input to this process. If we observe a
difference in categorization between East Asian and Western
participants, it could be that members of both groups use the same
procedure (considering similarity that is influenced by both taxonomic
and thematic relations), but the input to this procedure differ between
cultures, with East Asian participants exposed to more support for
thematic similarity in the language they hear in comparison to their
Western counterparts. We might also expect that both conceptualization
of similarity and input from experience play a role in driving
cross-cultural differences -- East Asian participants may be exposed to
more instances of thematic similarity, and prefer to conceptualize
similarity as thematic to a greater extent than Westerners.

@@

To determine whether varied input to similarity judgments can in part
explain cross-cultural differences, we need a way to operationalize
variation in exposure to thematic and taxonomic relations. While
exposure to types of similarity is difficult to measure, the statistics
of language can provide a rough proxy. Language statistics are useful in
that they are part of the input to everyday experience -- and indeed,
may afford many of the `experiences' that people have with infrequently
encountered items, like cows or helicopters -- and they provide an
accessible measure. Previous work also suggests that language
statistics, such as lexical co-occurrence or cosine distance of word
embeddings, can be good predictors of similarity reasoning. Semantic
models that are constructed using lexical co-occurrence (in comparison
to annotated relations) have been shown to perform well on predicting
human judgments about similarity between word pairs that are
thematically or taxonomically related (Rohde, Gonnerman, \& Plaut,
2006). Relatedly, a model trained on word-document co-occurrence can
predict word association and the effects of semantic association on a
variety of linguistic tasks (Griffiths, Steyvers, \& Tenenbaum, 2007).
Word embeddings like word2vec, gloVe, and fastText have also been shown
to be good predictors for similarity judgments Jatnika, Bijaksana, \&
Suryani (2019).

Our study uses cosine distances of fastText word vectors as a measure of
lexical co-occurrence\footnote{We also carried out our analysis using
  raw lexical co-occurrences and obtained similar results.}. fastText is
a system that uses lexical co-occurrence information to generate a
vector representing each word in its lexicon (Mikolov, Grave,
Bojanowski, Puhrsch, \& Joulin, 2018). fastText has also been shown to
be sensitive to cultural differences in word meanings: Thompson \&
Lupyan (2020) demonstrated that the distribution of semantic meaning
clusters generated by fastText trained on language-specific corpora
correlates with the cultural, historical, and geographical similarities
of these languages.

We note that language statistics may incorporate both cultural-specific
environmental statistics, that is, the experiential \emph{input} to
similarity judgments, and culture-specific senses of similarity, the
\emph{conceptualization} of similarity. For example, cultural features
(like farming) can lead to differences in environmental statistics
(seeing cows and grass) and these can influence language (talking more
about cows and grass). But other cultural features (like conceiving of
similarity thematically) could also cause individuals to talk
differently about the same experiences (mentioning what cows eat rather
than what other animals cows are like). Acoordingly, our approach
examines the extent to which language statistics can predict
cross-cultural differences in similarity judgments with the
understanding that language statistics are likely a proxy for both
\emph{input} to and \emph{conceptualization} of similarity.

@@

The present study tests whether the statistics of language can be used
to predict cross-cultural differences in similarity judgments, and
particularly, whether this approach provides additional insight beyond
cross-cultural characterizations based on taxonomic vs thematic
preferences.

We measured taxonomic versus thematic similarity matching in a
forced-choice word triad task in three populations. Following Ji et al.
(2004), we measured preferences in the US and China. In addition, we
collected data in a novel context: Vietnam. Vietnam is a Southeast Asian
country that borders China and has historically been greatly influenced
by Chinese culture (Hui, 2002). Therefore, it serves as a suitable
cultural context to investigate whether the claim made by Ji et al.
(2004) and previous studies -- that Eastern and Western cultures have
different notions of similarity -- extends beyond mainland China. In
addition to these replication and extension questions, we tested whether
fastText vectors from corpora corresponding to each language context
(English, Mandarin, and Vietnamese) are good predictors for similarity
judgments in each population.

This study is correlational and cannot evaluate causal relationships
between environmental statistics and similarity judgments. However, this
work can inform potential mechanisms by examining whether similarity
judgments covary with environmental statistics that differ across these
contexts. Our specific research questions are as follows:~

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Do we replicate cross-cultural differences in similarity judgments
  between East Asian and Western cultures?
\item
  Are these cross-cultural differences related to differences in
  language?
\item
  Is our language model specific to items with a taxonomic-thematic
  contrast, or can it predict similarity more generally?~
\end{enumerate}

To preview our results, we replicate US-China differences and find that
Vietnamese judgments are intermediate between these two. We find that
language-specific statistics provide good predictions for cross-cultural
differences in similarity judgments, and additionally, for more general
similarity judgments that do not contrast taxonomic and thematic
matches. Critically, the stimuli used to assess this more general case
of similarity reasoning were constructed to be outside the scope of
explanation for taxonomic-thematic accounts (for use as filler items).
While the language model may succeed in taxonomic-thematic similarity
predictions by picking up on differences in the \emph{conceptualization}
of similarity that are reflected in language data, this
conceptualization account provides no prediction for these filler
stimuli. Accordingly, these filler stimuli provide the strongest test of
the language model, demonstrating that it can explain variation in
similarity judgments beyond that explained by thematic or taxonomic
conceptualizations of similarity. Taken together, our findings provide
support for an alternative to previous accounts, on which differing
cultures induce differing conceptions of similarity. They show that
language can account for culture-specific variation in similarity
reasoning, both for taxonomic-thematic judgments and for judgments not
attributable to these senses of similarity. While these findings do not
rule out contributions from differing senses of similarity, they show it
is possible to explain cross-cultural differences in similarity
judgments without invoking variation in notions of similarity, at least
in some cases.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Nisbett2003b}{}}%
Nisbett, R. E., \& Masada, T. (2003). Culture and point of view.
\emph{Proceedings of the National Academy of Sciences}, \emph{100}(19),
11163--11170.
http://doi.org/\href{https://doi.org/10.1073/pnas.1934527100}{10.1073/pnas.1934527100}
